---
title: "01_clean_data"
author: "Jiarong Feng"
date: "2026-06-02"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
source("../utils/utils.R")
source("../utils/clean_data_utils.R")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

theme_set(theme_plotwise())
```

# Overview

This notebook loads the four raw condition files, adds condition labels, validates participant structure, preserves survey and open-text responses, separates major study phases, and saves clean datasets for next scoring and analysis.

This notebook is only for data cleaning

# 1. Raw Files and Study Constants

```{r raw-file-config}
raw_files <- tribble(
  ~filename,                         ~lecture_type, ~practice_type, ~group,
  "main-control-control.csv",         "Control",     "Control",      "CC",
  "main-control-plotwise.csv",        "Control",     "PlotWise",     "CP",
  "main-plotwise-control.csv",        "PlotWise",    "Control",      "PC",
  "main-plotwise-plotwise.csv",       "PlotWise",    "PlotWise",     "PP"
)

PRE_SURVEY_IDS <- c(
  "q-correlation",
  "q-slope",
  "q-class-taken",
  "q-how-many",
  "q-class-time",
  "q-class-level",
  "q-grade"
)

STATS_BACKGROUND_IDS <- c(
  "q-class-taken",
  "q-how-many",
  "q-class-time",
  "q-class-level",
  "q-grade"
)

PRACTICE_MOTIVATION_IDS <- c("q-motivation")

ASSESSMENT_VALUE_ESTIMATION_TRIALS <- c(
  paste0("guess-correlation-", 1:8),
  paste0("guess-slope-", 1:8)
)

ASSESSMENT_CHOOSE_R_TRIALS <- c(
  "choose-strongest-r-1",
  "choose-strongest-r-2"
)

ASSESSMENT_ADD_POINTS_TRIALS <- c(
  "add-points-1",
  "add-points-2",
  "add-points-3"
)

ASSESSMENT_CONNECTION_MAKING_TRIALS <- c("connection-making")

ASSESSMENT_CONNECTION_MAKING_IDS <- c(
  "switch-question-response-a",
  "switch-question-response-b",
  "q-same",
  "q-different"
)

INTEREST_TRIALS <- c(
  "triggeredSituationalInterest",
  "maintainedSituationalInterest",
  "futureInterestSurvey"
)
```

# 2. Load Raw Data

```{r load-raw-data}
raw_list <- pmap(
  raw_files,
  \(filename, lecture_type, practice_type, group) {
    read_group(
      filename = filename,
      lecture_type_value = lecture_type,
      practice_type_value = practice_type,
      group_value = group
    )
  }
)

names(raw_list) <- raw_files$group

raw_all <- bind_rows(raw_list) |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  )

cat("Loaded raw data:\n")
cat("- Total rows:", nrow(raw_all), "\n")
cat("- Total columns:", ncol(raw_all), "\n")
cat("- Unique participants:", n_distinct(raw_all$participantId), "\n")
```

# 3. Data Validation

```{r data-validation}
participant_counts <- raw_all |>
  distinct(participantId, group, lecture_type, practice_type) |>
  count(group, lecture_type, practice_type, name = "n_participants")

duplicate_participant_check <- raw_all |>
  distinct(participantId, group) |>
  count(participantId, name = "n_conditions") |>
  filter(n_conditions > 1)

row_counts <- raw_all |>
  count(group, source_file, name = "n_rows") |>
  arrange(group)

column_inventory <- tibble(
  column_name = names(raw_all),
  column_type = map_chr(raw_all, \(x) class(x)[1]),
  n_missing = map_int(raw_all, \(x) sum(is.na(x))),
  pct_missing = n_missing / nrow(raw_all)
) |>
  arrange(desc(pct_missing), column_name)

trial_overview <- raw_all |>
  count(group, trialId, name = "n_rows") |>
  arrange(group, trialId)

response_overview <- raw_all |>
  count(group, responseId, name = "n_rows") |>
  arrange(group, responseId)

trial_response_overview <- raw_all |>
  count(group, trialId, responseId, responsePrompt, name = "n_rows") |>
  arrange(group, trialId, responseId)

stopifnot(sum(participant_counts$n_participants) == 80)
stopifnot(all(participant_counts$n_participants == 20))
stopifnot(nrow(duplicate_participant_check) == 0)

cat("Validation passed:\n")
cat("- 80 unique participants\n")
cat("- 20 participants per condition\n")
cat("- no participant appears in more than one condition\n")
```

# 4. Participant Index

```{r participant-index}
participant_condition_index <- raw_all |>
  group_by(participantId, group, lecture_type, practice_type) |>
  summarise(
    status = first(na.omit(status)),
    percentComplete = max(percentComplete, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(group, participantId)

participant_completion_summary <- participant_condition_index |>
  count(group, status, percentComplete, name = "n") |>
  arrange(group, status, percentComplete)

cat("Participant index created:", nrow(participant_condition_index), "participants\n")
```

# 5. Item-Level Dataset

The item-level dataset keeps all original columns and adds condition variables. This is the main dataset used for analysis.

```{r item-level}
clean_item_level <- raw_all |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  ) |>
  arrange(group, participantId, trialId, responseId)

cat("Clean item-level dataset created:\n")
cat("- Rows:", nrow(clean_item_level), "\n")
cat("- Columns:", ncol(clean_item_level), "\n")
```

# 6. Save All Non-Empty Answer Rows

This file save every non-empty response in a simple long-format table. Might be useful for future analysis.

```{r all-answer-rows}
all_answer_rows <- clean_item_level |>
  filter(is_nonempty_text(answer)) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "source_file",
      "trialId",
      "responseId",
      "responsePrompt",
      "answer",
      "duration",
      "cleanedDuration",
      "status",
      "percentComplete"
    )),
    everything()
  ) |>
  arrange(group, participantId, trialId, responseId)

cat("All non-empty answer rows:", nrow(all_answer_rows), "\n")
```

# 7. Pre-Survey Data


```{r pre-survey}
pre_survey_long <- clean_item_level |>
  filter(
    trialId == "pre questions",
    responseId %in% PRE_SURVEY_IDS
  ) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "source_file",
      "trialId",
      "responseId",
      "responsePrompt",
      "answer",
      "duration",
      "cleanedDuration",
      "status",
      "percentComplete"
    ))
  ) |>
  arrange(group, participantId, responseId)

pre_survey_wide <- pre_survey_long |>
  group_by(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId
  ) |>
  summarise(
    answer = first_non_missing(answer),
    responsePrompt = first_non_missing(responsePrompt),
    .groups = "drop"
  ) |>
  select(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId,
    answer
  ) |>
  pivot_wider(
    names_from = responseId,
    values_from = answer,
    names_prefix = "pre_"
  ) |>
  arrange(group, participantId)

cat("Pre-survey rows:", nrow(pre_survey_long), "\n")
```

# 8. Definition Responses for Prior Knowledge Coding

These are the pre-survey definition responses that will later be coded as prior knowledge.

```{r definition-responses}
definition_responses <- pre_survey_long |>
  filter(responseId %in% c("q-correlation", "q-slope")) |>
  filter(is_nonempty_text(answer)) |>
  arrange(group, participantId, responseId)

definition_responses_wide <- definition_responses |>
  group_by(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId
  ) |>
  summarise(
    answer = first_non_missing(answer),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = responseId,
    values_from = answer,
    names_prefix = "definition_"
  ) |>
  arrange(group, participantId)

cat("Definition response rows:", nrow(definition_responses), "\n")
```

# 9. Statistics Background Responses

These fields support later prior-knowledge analysis based on self-reported statistics background.

```{r stats-background}
stats_background_long <- pre_survey_long |>
  filter(responseId %in% STATS_BACKGROUND_IDS) |>
  arrange(group, participantId, responseId)

stats_background_wide <- stats_background_long |>
  group_by(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId
  ) |>
  summarise(
    answer = first_non_missing(answer),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = responseId,
    values_from = answer,
    names_prefix = "stats_"
  ) |>
  arrange(group, participantId)

stats_background_distribution <- stats_background_long |>
  count(responseId, responsePrompt, answer, name = "n_rows") |>
  arrange(responseId, answer)

cat("Statistics background rows:", nrow(stats_background_long), "\n")
```

# 10. Practice Motivation Response

`q-motivation` appears after practice and before assessment. It captures whether participants wanted to practice more, so it is preserved separately from pre-survey and interest scales.

```{r practice-motivation}
practice_motivation_long <- clean_item_level |>
  filter(responseId %in% PRACTICE_MOTIVATION_IDS) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "source_file",
      "trialId",
      "responseId",
      "responsePrompt",
      "answer",
      "duration",
      "cleanedDuration",
      "status",
      "percentComplete"
    ))
  ) |>
  arrange(group, participantId, trialId, responseId)

practice_motivation_wide <- practice_motivation_long |>
  group_by(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId
  ) |>
  summarise(
    answer = first_non_missing(answer),
    responsePrompt = first_non_missing(responsePrompt),
    .groups = "drop"
  ) |>
  select(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId,
    answer
  ) |>
  pivot_wider(
    names_from = responseId,
    values_from = answer,
    names_prefix = "practice_"
  ) |>
  arrange(group, participantId)

cat("Practice motivation rows:", nrow(practice_motivation_long), "\n")
```

# 11. Assessment Data Subsets

This section preserves assessment components separately without scoring them.

## 11.1 Value Estimation Assessment

```{r assessment-value-estimation}
assessment_value_estimation_long <- clean_item_level |>
  filter(trialId %in% ASSESSMENT_VALUE_ESTIMATION_TRIALS) |>
  arrange(group, participantId, trialId, responseId)

cat("Assessment value-estimation rows:", nrow(assessment_value_estimation_long), "\n")
```

## 11.2 Choose Stronger Correlation Assessment

```{r assessment-choose-r}
assessment_choose_r_long <- clean_item_level |>
  filter(trialId %in% ASSESSMENT_CHOOSE_R_TRIALS) |>
  arrange(group, participantId, trialId, responseId)

cat("Assessment choose-r rows:", nrow(assessment_choose_r_long), "\n")
```

## 11.3 Add-Points Assessment

```{r assessment-add-points}
assessment_add_points_long <- clean_item_level |>
  filter(trialId %in% ASSESSMENT_ADD_POINTS_TRIALS) |>
  arrange(group, participantId, trialId, responseId)

cat("Assessment add-points rows:", nrow(assessment_add_points_long), "\n")
```

## 11.4 Connection-Making Assessment

The connection-making questions are open response.

```{r assessment-connection-making}
assessment_connection_making_long <- clean_item_level |>
  filter(
    trialId %in% ASSESSMENT_CONNECTION_MAKING_TRIALS,
    responseId %in% ASSESSMENT_CONNECTION_MAKING_IDS
  ) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "source_file",
      "trialId",
      "responseId",
      "responsePrompt",
      "answer",
      "duration",
      "cleanedDuration",
      "status",
      "percentComplete"
    ))
  ) |>
  arrange(group, participantId, responseId)

assessment_connection_making_wide <- assessment_connection_making_long |>
  group_by(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId
  ) |>
  summarise(
    answer = first_non_missing(answer),
    responsePrompt = first_non_missing(responsePrompt),
    .groups = "drop"
  ) |>
  select(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId,
    answer
  ) |>
  pivot_wider(
    names_from = responseId,
    values_from = answer,
    names_prefix = "assessment_connection_"
  ) |>
  arrange(group, participantId)

cat("Assessment connection-making rows:", nrow(assessment_connection_making_long), "\n")
```

# 12. Interest Survey Data

This section only preserves raw interest-survey rows. Reverse-coding will be done in 06 notebook

```{r interest-survey}
interest_survey_long <- clean_item_level |>
  filter(trialId %in% INTEREST_TRIALS) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "source_file",
      "trialId",
      "responseId",
      "responsePrompt",
      "answer",
      "duration",
      "cleanedDuration",
      "status",
      "percentComplete"
    ))
  ) |>
  arrange(group, participantId, trialId, responseId)

interest_survey_wide <- interest_survey_long |>
  group_by(
    participantId,
    group,
    lecture_type,
    practice_type,
    trialId,
    responseId
  ) |>
  summarise(
    answer = first_non_missing(answer),
    responsePrompt = first_non_missing(responsePrompt),
    .groups = "drop"
  ) |>
  select(
    participantId,
    group,
    lecture_type,
    practice_type,
    trialId,
    responseId,
    answer
  ) |>
  mutate(item_id = paste(trialId, responseId, sep = "__")) |>
  select(-trialId, -responseId) |>
  pivot_wider(
    names_from = item_id,
    values_from = answer,
    names_prefix = "interest_"
  ) |>
  arrange(group, participantId)

cat("Interest survey rows:", nrow(interest_survey_long), "\n")
```

# 13. Open-Text responses

This broad file preserves likely open-text responses across phases. It is intentionally inclusive and is not used to define any specific study phase.

```{r open-text-candidates}
all_open_text_candidates <- clean_item_level |>
  mutate(
    answer_chr = as.character(answer),
    answer_trimmed = str_trim(answer_chr),
    answer_num = suppressWarnings(as.numeric(answer_trimmed)),
    answer_length = str_length(answer_trimmed),
    looks_like_json = str_detect(answer_trimmed, "^\\s*[\\[{]"),
    looks_numeric = !is.na(answer_num),
    likely_open_text = case_when(
      responseId %in% c(
        "q-correlation",
        "q-slope",
        "q-motivation",
        "switch-question-response-a",
        "switch-question-response-b",
        "q-same",
        "q-different"
      ) ~ TRUE,
      str_detect(
        str_to_lower(coalesce(responsePrompt, "")),
        "describe|explain|list|why|how|question|comment|thought|similar|different|motivation"
      ) ~ TRUE,
      answer_length >= 25 & !looks_like_json & !looks_numeric ~ TRUE,
      TRUE ~ FALSE
    )
  ) |>
  filter(
    is_nonempty_text(answer_trimmed),
    likely_open_text
  ) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "source_file",
      "trialId",
      "responseId",
      "responsePrompt",
      "answer",
      "duration",
      "cleanedDuration",
      "status",
      "percentComplete"
    )),
    answer_length
  ) |>
  arrange(group, participantId, trialId, responseId)

cat("Open-text candidate rows:", nrow(all_open_text_candidates), "\n")
```

# 14. Save Clean Outputs

```{r save-clean-outputs}
write_csv(clean_item_level, file.path(data_derived_dir, "clean_item_level.csv"))
write_csv(participant_condition_index, file.path(data_derived_dir, "participant_condition_index.csv"))
write_csv(all_answer_rows, file.path(data_derived_dir, "all_answer_rows.csv"))

write_csv(pre_survey_long, file.path(data_derived_dir, "pre_survey_long.csv"))
write_csv(pre_survey_wide, file.path(data_derived_dir, "pre_survey_wide.csv"))
write_csv(definition_responses, file.path(data_derived_dir, "definition_responses.csv"))
write_csv(definition_responses_wide, file.path(data_derived_dir, "definition_responses_wide.csv"))
write_csv(stats_background_long, file.path(data_derived_dir, "stats_background_long.csv"))
write_csv(stats_background_wide, file.path(data_derived_dir, "stats_background_wide.csv"))

write_csv(practice_motivation_long, file.path(data_derived_dir, "practice_motivation_long.csv"))
write_csv(practice_motivation_wide, file.path(data_derived_dir, "practice_motivation_wide.csv"))

write_csv(assessment_value_estimation_long, file.path(data_derived_dir, "assessment_value_estimation_long.csv"))
write_csv(assessment_choose_r_long, file.path(data_derived_dir, "assessment_choose_r_long.csv"))
write_csv(assessment_add_points_long, file.path(data_derived_dir, "assessment_add_points_long.csv"))
write_csv(assessment_connection_making_long, file.path(data_derived_dir, "assessment_connection_making_long.csv"))
write_csv(assessment_connection_making_wide, file.path(data_derived_dir, "assessment_connection_making_wide.csv"))

write_csv(interest_survey_long, file.path(data_derived_dir, "interest_survey_long.csv"))
write_csv(interest_survey_wide, file.path(data_derived_dir, "interest_survey_wide.csv"))

write_csv(all_open_text_candidates, file.path(data_derived_dir, "all_open_text_candidates.csv"))

# These are only for debugging
write_csv(row_counts, file.path(output_dir, "row_counts.csv"))
write_csv(column_inventory, file.path(output_dir, "column_inventory.csv"))
write_csv(participant_counts, file.path(output_dir, "participant_counts.csv"))
write_csv(participant_completion_summary, file.path(output_dir, "participant_completion_summary.csv"))
write_csv(trial_overview, file.path(output_dir, "trial_overview.csv"))
write_csv(response_overview, file.path(output_dir, "response_overview.csv"))
write_csv(trial_response_overview, file.path(output_dir, "trial_response_overview.csv"))
write_csv(stats_background_distribution, file.path(output_dir, "stats_background_distribution.csv"))

cat("Saved clean data outputs to:", data_derived_dir, "\n")
cat("Saved overview outputs to:", output_dir, "\n")
```

# 15. Output Files

```{r output-files}
created_files <- tibble(
  file = c(
    "clean_item_level.csv",
    "participant_condition_index.csv",
    "all_answer_rows.csv",
    "pre_survey_long.csv",
    "pre_survey_wide.csv",
    "definition_responses.csv",
    "definition_responses_wide.csv",
    "stats_background_long.csv",
    "stats_background_wide.csv",
    "practice_motivation_long.csv",
    "practice_motivation_wide.csv",
    "assessment_value_estimation_long.csv",
    "assessment_choose_r_long.csv",
    "assessment_add_points_long.csv",
    "assessment_connection_making_long.csv",
    "assessment_connection_making_wide.csv",
    "interest_survey_long.csv",
    "interest_survey_wide.csv",
    "all_open_text_candidates.csv"
  ),
  folder = data_derived_dir
)

created_overview_files <- tibble(
  file = c(
    "row_counts.csv",
    "column_inventory.csv",
    "participant_counts.csv",
    "participant_completion_summary.csv",
    "trial_overview.csv",
    "response_overview.csv",
    "trial_response_overview.csv",
    "stats_background_distribution.csv"
  ),
  folder = output_dir
)

cat("Created", nrow(created_files), "derived data files and",
    nrow(created_overview_files), "overview files.\n")
```