---
title: "02_parse_and_score_outcomes"
author: "Jiarong Feng"
date: "2026-06-02"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
source("../utils/utils.R")
source("../utils/scoring_utils.R")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

theme_set(theme_plotwise())
```

# Overview

This notebook parses the cleaned response data created by `01_clean_data.Rmd`.

The goal is to convert raw responses into item-level and participant-level outcomes for later analysis.

# 1. Scoring Constants

```{r scoring-constants}
# These values are the initial r and m shown in the add-points assessment.
INITIAL_R <- 0.4211
INITIAL_SLOPE <- 0.3156
```


# 2. Load Cleaned Data

```{r load-clean-data}
participant_condition_index <- read_csv(
  file.path(data_derived_dir, "participant_condition_index.csv"),
  show_col_types = FALSE
) |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  )

assessment_value_estimation_long <- read_csv(
  file.path(data_derived_dir, "assessment_value_estimation_long.csv"),
  show_col_types = FALSE
) |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  )

assessment_choose_r_long <- read_csv(
  file.path(data_derived_dir, "assessment_choose_r_long.csv"),
  show_col_types = FALSE
) |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  )

assessment_add_points_long <- read_csv(
  file.path(data_derived_dir, "assessment_add_points_long.csv"),
  show_col_types = FALSE
) |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  )

interest_survey_long <- read_csv(
  file.path(data_derived_dir, "interest_survey_long.csv"),
  show_col_types = FALSE
) |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  )

cat("Loaded cleaned datasets.\n")
```

# 4. Score Value Estimation Assessment

Value-estimation items ask participants to estimate either correlation or slope. We score these using absolute error from the correct value.

Extreme slope entries were flagged when `|answer| > 10`. We preserved the original responses and original absolute errors, but set these flagged entries to `NA` in the primary cleaned error variable (`abs_error_no_extreme`). No decimal correction was applied.

```{r score-value-estimation-item-level}
scored_value_estimation_item_level <- assessment_value_estimation_long |>
  filter(responseId == "stat_guess") |>
  mutate(
    stat_type = case_when(
      str_detect(trialId, "correlation") ~ "correlation",
      str_detect(trialId, "slope") ~ "slope",
      TRUE ~ NA_character_
    ),
    answer_num = suppressWarnings(as.numeric(answer)),
    correct_answer = map2_dbl(parameters_imgLabel, stat_type, parse_correct_answer),
    abs_error = abs(answer_num - correct_answer),
    value_sign = case_when(
      correct_answer > 0 ~ "positive",
      correct_answer < 0 ~ "negative",
      correct_answer == 0 ~ "zero",
      TRUE ~ NA_character_
    ),
    sign_error = case_when(
      is.na(answer_num) | is.na(correct_answer) ~ NA,
      correct_answer == 0 ~ FALSE,
      answer_num * correct_answer < 0 ~ TRUE,
      TRUE ~ FALSE
    ),
    extreme_entry = case_when(
      stat_type == "slope" & !is.na(answer_num) & abs(answer_num) > 10 ~ TRUE,
      TRUE ~ FALSE
    ),
    abs_error_no_extreme = if_else(extreme_entry, NA_real_, abs_error)
  ) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "trialId",
      "responseId",
      "stat_type",
      "answer",
      "answer_num",
      "correct_answer",
      "abs_error",
      "abs_error_no_extreme",
      "extreme_entry",
      "value_sign",
      "sign_error",
      "duration",
      "cleanedDuration",
      "parameters_imgLabel"
    ))
  ) |>
  arrange(group, participantId, stat_type, trialId)

cat("Scored value-estimation item rows:", nrow(scored_value_estimation_item_level), "\n")
cat("Extreme slope entries:", sum(scored_value_estimation_item_level$extreme_entry, na.rm = TRUE), "\n")
```


```{r review-extreme-value-estimation-entries}
extreme_value_estimation_entries <- scored_value_estimation_item_level |>
  filter(extreme_entry) |>
  select(
    participantId,
    group,
    lecture_type,
    practice_type,
    trialId,
    stat_type,
    answer,
    answer_num,
    correct_answer,
    abs_error,
    parameters_imgLabel,
    duration,
    cleanedDuration
  ) |>
  arrange(group, participantId, trialId)

cat("Extreme value-estimation entries:", nrow(extreme_value_estimation_entries), "\n")

extreme_value_estimation_entries

extreme_value_estimation_entries |>
  count(group, participantId, name = "n_extreme_entries") |>
  arrange(group, desc(n_extreme_entries), participantId)

extreme_value_estimation_entries |>
  count(group, trialId, correct_answer, name = "n_extreme_entries") |>
  arrange(group, trialId)
```


```{r score-value-estimation-participant-level}
scored_value_estimation_participant_level <- scored_value_estimation_item_level |>
  group_by(participantId, group, lecture_type, practice_type, stat_type) |>
  summarise(
    n_items = n(),
    n_valid = sum(!is.na(abs_error_no_extreme)),
    n_extreme = sum(extreme_entry, na.rm = TRUE),
    mean_abs_error = mean(abs_error_no_extreme, na.rm = TRUE),
    median_abs_error = median(abs_error_no_extreme, na.rm = TRUE),
    mean_abs_error_all = mean(abs_error, na.rm = TRUE),
    median_abs_error_all = median(abs_error, na.rm = TRUE),
    sign_error_rate = mean(as.numeric(sign_error), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(group, participantId, stat_type)

cat("Scored value-estimation participant rows:", nrow(scored_value_estimation_participant_level), "\n")
```

# 5. Score Choose-Strongest-Correlation Assessment

These items ask participants to choose which graph has the stronger correlation. We compare the absolute correlation shown in each graph label.

```{r score-choose-r-item-level}
scored_choose_r_item_level <- assessment_choose_r_long |>
  filter(responseId == "graph_choice") |>
  mutate(
    abs_r_left = map_dbl(parameters_leftLabel, parse_r_from_label),
    abs_r_right = map_dbl(parameters_rightLabel, parse_r_from_label),
    correct_side = case_when(
      abs_r_left > abs_r_right ~ "left",
      abs_r_right > abs_r_left ~ "right",
      TRUE ~ NA_character_
    ),
    answer_clean = str_to_lower(str_trim(answer)),
    is_correct = case_when(
      is.na(correct_side) | is.na(answer_clean) ~ NA,
      answer_clean == correct_side ~ TRUE,
      TRUE ~ FALSE
    )
  ) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "trialId",
      "responseId",
      "answer",
      "answer_clean",
      "correct_side",
      "is_correct",
      "abs_r_left",
      "abs_r_right",
      "duration",
      "cleanedDuration",
      "parameters_leftLabel",
      "parameters_rightLabel"
    ))
  ) |>
  arrange(group, participantId, trialId)

cat("Scored choose-r item rows:", nrow(scored_choose_r_item_level), "\n")
```

```{r score-choose-r-participant-level}
scored_choose_r_participant_level <- scored_choose_r_item_level |>
  group_by(participantId, group, lecture_type, practice_type) |>
  summarise(
    choose_r_n_items = n(),
    choose_r_score = sum(is_correct, na.rm = TRUE),
    choose_r_pct_correct = mean(as.numeric(is_correct), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(group, participantId)

cat("Scored choose-r participant rows:", nrow(scored_choose_r_participant_level), "\n")
```

# 6. Score Add-Points Assessment

The add-points assessment asks participants to change a scatterplot by adding points while trying to not change the slope.

We score several outcomes:

- `r_correct_direction`: whether final correlation moved in the intended direction.
- `intended_delta_r`: how much final correlation moved in the intended direction.
- `reached_negative_r`: for add-points-2 only, whether the final correlation crossed below zero.
- `abs_delta_slope`: how much the slope changed from the initial plot.

Scoring details:

- `add-points-1`: increase positive correlation.
- `add-points-2`: increase negative correlation. We score partial success as `final_r < initial_r`, and separately retain the stricter indicator `final_r < 0`.
- `add-points-3`: move correlation closer to zero, scored using absolute distance from zero.

```{r score-add-points-item-level}
scored_add_points_item_level <- assessment_add_points_long |>
  filter(responseId == "add_points") |>
  mutate(
    stats = map(answer, extract_add_points_stats),

    n_points = map_int(stats, "n_points"),
    final_r = map_dbl(stats, "final_r"),
    final_slope = map_dbl(stats, "final_slope"),

    initial_r = INITIAL_R,
    initial_slope = INITIAL_SLOPE,

    delta_r = final_r - initial_r,
    abs_delta_r = abs(delta_r),

    delta_slope = final_slope - initial_slope,
    abs_delta_slope = abs(delta_slope),

    task_goal = case_when(
      trialId == "add-points-1" ~ "Increase positive correlation",
      trialId == "add-points-2" ~ "Increase negative correlation",
      trialId == "add-points-3" ~ "Move correlation closer to zero",
      TRUE ~ trialId
    ),

    r_correct_direction = case_when(
      trialId == "add-points-1" ~ final_r > initial_r,
      trialId == "add-points-2" ~ final_r < initial_r,
      trialId == "add-points-3" ~ abs(final_r) < abs(initial_r),
      TRUE ~ NA
    ),

    intended_delta_r = case_when(
      trialId == "add-points-1" ~ final_r - initial_r,
      trialId == "add-points-2" ~ initial_r - final_r,
      trialId == "add-points-3" ~ abs(initial_r) - abs(final_r),
      TRUE ~ NA_real_
    ),

    reached_negative_r = case_when(
      trialId == "add-points-2" ~ final_r < 0,
      TRUE ~ NA
    )
  ) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "trialId",
      "responseId",
      "task_goal",
      "answer",
      "n_points",
      "initial_r",
      "initial_slope",
      "final_r",
      "final_slope",
      "delta_r",
      "abs_delta_r",
      "intended_delta_r",
      "delta_slope",
      "abs_delta_slope",
      "r_correct_direction",
      "reached_negative_r",
      "duration",
      "cleanedDuration"
    ))
  ) |>
  arrange(group, participantId, trialId)

cat("Scored add-points item rows:", nrow(scored_add_points_item_level), "\n")
```

```{r score-add-points-participant-level}
scored_add_points_participant_level <- scored_add_points_item_level |>
  group_by(participantId, group, lecture_type, practice_type) |>
  summarise(
    add_points_n_items = n(),
    add_points_mean_n_points = mean(n_points, na.rm = TRUE),
    add_points_pct_r_correct = mean(as.numeric(r_correct_direction), na.rm = TRUE),
    add_points_mean_intended_delta_r = mean(intended_delta_r, na.rm = TRUE),
    add_points_mean_final_r = mean(final_r, na.rm = TRUE),
    add_points_mean_delta_r = mean(delta_r, na.rm = TRUE),
    add_points_mean_abs_delta_r = mean(abs_delta_r, na.rm = TRUE),
    add_points_mean_final_slope = mean(final_slope, na.rm = TRUE),
    add_points_mean_delta_slope = mean(delta_slope, na.rm = TRUE),
    add_points_mean_abs_delta_slope = mean(abs_delta_slope, na.rm = TRUE),
    add_points_pct_reached_negative_r = mean(as.numeric(reached_negative_r), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(group, participantId)

cat("Scored add-points participant rows:", nrow(scored_add_points_participant_level), "\n")
```

# 7. Score Interest Survey

Interest items are scored on a 1--6 scale. Reverse-coded items are recoded as `7 - rating`.

```{r interest-scale-metadata}
interest_scale_meta <- tribble(
  ~responseId, ~scale_type, ~question_label, ~reverse,

  "tsi_fun_video", "TSI", "Video was fun", FALSE,
  "tsi_boring_video", "TSI", "Video was boring", TRUE,
  "tsi_enjoyed_session", "TSI", "Enjoyed this session", FALSE,
  "tsi_session_interesting", "TSI", "Session was interesting", FALSE,
  "tsi_enjoyed_learning", "TSI", "Enjoyed learning new things", FALSE,
  "tsi_session_fun", "TSI", "Session was fun", FALSE,
  "tsi_session_boring", "TSI", "Session was boring", TRUE,
  "tsi_fun_stats", "TSI", "Fun to learn about statistics this way", FALSE,

  "msi_excited", "MSI", "Excited about correlation and slope", FALSE,
  "msi_interesting_corrslope", "MSI", "Correlation and slope are interesting", FALSE,
  "msi_interesting_stats", "MSI", "Statistics is interesting", FALSE,
  "msi_enjoyed_learning", "MSI", "Enjoyed learning about correlation and slope", FALSE,

  "interest_any_study", "FutureInterest", "Interested in any study", FALSE,
  "interest_statistics_study", "FutureInterest", "Interested in a statistics study", FALSE,
  "interest_dynamic_vis_study", "FutureInterest", "Interested in a dynamic visualization study", FALSE
)
```

```{r score-interest-item-level}
scored_interest_item_level <- interest_survey_long |>
  filter(responseId %in% interest_scale_meta$responseId) |>
  mutate(
    answer_num = suppressWarnings(as.numeric(answer))
  ) |>
  left_join(interest_scale_meta, by = "responseId") |>
  mutate(
    answer_scored = if_else(reverse, 7 - answer_num, answer_num)
  ) |>
  select(
    any_of(c(
      "participantId",
      "group",
      "lecture_type",
      "practice_type",
      "trialId",
      "responseId",
      "responsePrompt",
      "answer",
      "answer_num",
      "answer_scored",
      "scale_type",
      "question_label",
      "reverse",
      "duration",
      "cleanedDuration"
    ))
  ) |>
  arrange(group, participantId, scale_type, responseId)

cat("Scored interest item rows:", nrow(scored_interest_item_level), "\n")
```

```{r score-interest-participant-level}
scored_interest_participant_level <- scored_interest_item_level |>
  group_by(participantId, group, lecture_type, practice_type, scale_type) |>
  summarise(
    interest_n_items = n(),
    composite = mean(answer_scored, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(group, participantId, scale_type)

cat("Scored interest participant rows:", nrow(scored_interest_participant_level), "\n")
```

# 8. Build Participant-Level Outcome Table

This table has one row per participant and is the main input for later participant-level plots, bootstrap analyses, and summary tables.

```{r participant-outcomes}
value_estimation_wide <- scored_value_estimation_participant_level |>
  select(
    participantId,
    group,
    lecture_type,
    practice_type,
    stat_type,
    n_items,
    n_valid,
    n_extreme,
    mean_abs_error,
    median_abs_error,
    mean_abs_error_all,
    median_abs_error_all,
    sign_error_rate
  ) |>
  pivot_wider(
    names_from = stat_type,
    values_from = c(
      n_items,
      n_valid,
      n_extreme,
      mean_abs_error,
      median_abs_error,
      mean_abs_error_all,
      median_abs_error_all,
      sign_error_rate
    ),
    names_glue = "{stat_type}_{.value}"
  )

interest_wide <- scored_interest_participant_level |>
  select(
    participantId,
    group,
    lecture_type,
    practice_type,
    scale_type,
    composite
  ) |>
  pivot_wider(
    names_from = scale_type,
    values_from = composite
  )

participant_outcomes <- participant_condition_index |>
  select(participantId, group, lecture_type, practice_type, status, percentComplete) |>
  left_join(
    value_estimation_wide,
    by = c("participantId", "group", "lecture_type", "practice_type")
  ) |>
  left_join(
    scored_choose_r_participant_level,
    by = c("participantId", "group", "lecture_type", "practice_type")
  ) |>
  left_join(
    scored_add_points_participant_level,
    by = c("participantId", "group", "lecture_type", "practice_type")
  ) |>
  left_join(
    interest_wide,
    by = c("participantId", "group", "lecture_type", "practice_type")
  ) |>
  arrange(group, participantId)

cat("Participant outcome rows:", nrow(participant_outcomes), "\n")
cat("Participant outcome columns:", ncol(participant_outcomes), "\n")

stopifnot(nrow(participant_outcomes) == 80)
```

# 9. Save Scored Outputs

```{r save-scored-outputs}
write_csv(
  scored_value_estimation_item_level,
  file.path(data_derived_dir, "scored_value_estimation_item_level.csv")
)

write_csv(
  scored_value_estimation_participant_level,
  file.path(data_derived_dir, "scored_value_estimation_participant_level.csv")
)

write_csv(
  scored_choose_r_item_level,
  file.path(data_derived_dir, "scored_choose_r_item_level.csv")
)

write_csv(
  scored_choose_r_participant_level,
  file.path(data_derived_dir, "scored_choose_r_participant_level.csv")
)

write_csv(
  scored_add_points_item_level,
  file.path(data_derived_dir, "scored_add_points_item_level.csv")
)

write_csv(
  scored_add_points_participant_level,
  file.path(data_derived_dir, "scored_add_points_participant_level.csv")
)

write_csv(
  scored_interest_item_level,
  file.path(data_derived_dir, "scored_interest_item_level.csv")
)

write_csv(
  scored_interest_participant_level,
  file.path(data_derived_dir, "scored_interest_participant_level.csv")
)

write_csv(
  participant_outcomes,
  file.path(data_derived_dir, "participant_outcomes.csv")
)

cat("Saved scored outputs to:", data_derived_dir, "\n")
```

# 10. Output Files

```{r output-files}
created_files <- tibble(
  file = c(
    "scored_value_estimation_item_level.csv",
    "scored_value_estimation_participant_level.csv",
    "scored_choose_r_item_level.csv",
    "scored_choose_r_participant_level.csv",
    "scored_add_points_item_level.csv",
    "scored_add_points_participant_level.csv",
    "scored_interest_item_level.csv",
    "scored_interest_participant_level.csv",
    "participant_outcomes.csv"
  ),
  folder = data_derived_dir
)

cat("Created", nrow(created_files), "scored data files.\n")
```
