---
title: "03_prior_knowledge_coding"
author: "Jiarong Feng"
date: "2026-06-03"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
source("../utils/utils.R")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

theme_set(theme_plotwise())
```

# Overview

This notebook prepares and scores prior-knowledge responses from the pre-survey definition questions.

The prior-knowledge score is based on participants' open-ended definitions of:

1. correlation
2. slope

The goal is to create a participant-level file:

```text
data/derived/prior_knowledge_scores.csv
```

This file will be used in later analysis notebooks as a covariate:

```r
outcome ~ prior_knowledge_z + lecture_type * practice_type
```

# 1. Load Definition Responses

```{r load-definition-responses}
definition_responses <- read_csv(
  file.path(data_derived_dir, "definition_responses.csv"),
  show_col_types = FALSE
) |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  )

definition_responses_wide <- read_csv(
  file.path(data_derived_dir, "definition_responses_wide.csv"),
  show_col_types = FALSE
) |>
  mutate(
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    group = factor(group, levels = GROUP_LEVELS)
  )

cat("Definition response rows:", nrow(definition_responses), "\n")
cat("Definition response participants:", n_distinct(definition_responses$participantId), "\n")

definition_responses |>
  count(responseId, name = "n_rows")
```

# 2. Coding Rubric

Following the meeting discussion, we use a simple 0 / 0.5 / 1 rubric for each definition response.

Each participant can receive:

- 0 to 1 point for the correlation definition
- 0 to 1 point for the slope definition

The combined prior-knowledge score ranges from 0 to 2.

## 2.1 Correlation Definition Rubric

```text
1 = Mostly correct understanding
    - Defines correlation as a relationship/association between two variables
    - Or describes how two variables move/change together
    - May mention strength and/or direction of the relationship
    - Does not define correlation primarily as causation

0.5 = Partially correct understanding
    - Mentions a relationship, connection, association, or relatedness, but is vague
    - Refers to "things," "numbers," or "data sets" without clearly describing variables moving together
    - Shows some intuitive understanding but misses key ideas such as co-variation, strength, or direction
    - May include some causal language, but still indicates a relationship between variables

0 = Incorrect or no meaningful understanding
    - Blank, irrelevant, or "I don't know"
    - Defines an unrelated concept
    - Gives a very generic statement such as "something related somehow" without a meaningful statistical relationship
    - Strongly confuses correlation with an unrelated idea
```

## 2.2 Slope Definition Rubric

```text
1 = Mostly correct understanding
    - Defines slope as rate of change
    - Or describes change in y for a change in x
    - Or describes slope as rise over run
    - Or describes slope as the steepness of a line, with or without direction

0.5 = Partially correct understanding
    - Mentions direction, trend, incline, angle, or change, but does not clearly define steepness or rate of change
    - Mentions change in y only, without change relative to x
    - Refers to a line or two points on a graph, but does not clearly describe rise/run, steepness, or rate of change
    - Shows some intuitive understanding but is incomplete

0 = Incorrect or no meaningful understanding
    - Blank, irrelevant, or "I don't know"
    - Defines an unrelated concept
    - Confuses slope with distance between points
    - Strongly incorrect explanation
```

# 3. Create Coding Sheet

The coded file is `prior_knowledge_coding_sheet_coded.csv`.

```{r create-coding-sheet}
prior_knowledge_coding_sheet <- definition_responses |>
  transmute(
    participantId,
    group,
    lecture_type,
    practice_type,
    responseId,
    concept = case_when(
      responseId == "q-correlation" ~ "correlation",
      responseId == "q-slope" ~ "slope",
      TRUE ~ responseId
    ),
    responsePrompt,
    answer,
    score = NA_real_,
    coder_notes = NA_character_
  ) |>
  arrange(group, participantId, concept)

write_csv(
  prior_knowledge_coding_sheet,
  file.path(data_derived_dir, "prior_knowledge_coding_sheet.csv")
)

cat("Saved blank coding sheet to:",
    file.path(data_derived_dir, "prior_knowledge_coding_sheet.csv"),
    "\n")
```

# 4. Load Coded Prior-Knowledge Sheet

```{r load-coded-sheet}
coded_path <- file.path(data_derived_dir, "prior_knowledge_coding_sheet_coded.csv")

coded_prior_knowledge_available <- file.exists(coded_path)

if (coded_prior_knowledge_available) {
  prior_knowledge_coded <- read_csv(
    coded_path,
    show_col_types = FALSE
  ) |>
    mutate(
      lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
      practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
      group = factor(group, levels = GROUP_LEVELS),
      score = as.numeric(score)
    )

  cat("Loaded coded prior-knowledge sheet.\n")
  cat("Rows:", nrow(prior_knowledge_coded), "\n")
  cat("Participants:", n_distinct(prior_knowledge_coded$participantId), "\n")

  prior_knowledge_coded |>
    count(concept, score, name = "n")
} else {
  cat("No coded prior-knowledge sheet found yet.\n")
  cat("Please manually code prior_knowledge_coding_sheet.csv and save it as prior_knowledge_coding_sheet_coded.csv.\n")
}
```

# 5. Build Prior-Knowledge Scores


```{r build-prior-knowledge-scores}
if (coded_prior_knowledge_available) {
  prior_knowledge_scores <- prior_knowledge_coded |>
    group_by(participantId, group, lecture_type, practice_type) |>
    summarise(
      n_coded_items = sum(!is.na(score)),
      prior_knowledge_score = sum(score, na.rm = TRUE),
      prior_knowledge_mean = mean(score, na.rm = TRUE),
      correlation_definition_score = score[concept == "correlation"][1],
      slope_definition_score = score[concept == "slope"][1],
      .groups = "drop"
    ) |>
    mutate(
      prior_knowledge_z = as.numeric(scale(prior_knowledge_score))
    ) |>
    arrange(group, participantId)

  prior_knowledge_scores

  cat("Created prior-knowledge scores for",
      nrow(prior_knowledge_scores),
      "participants.\n")

  stopifnot(nrow(prior_knowledge_scores) == 80)
} else {
  cat("Skipped: coded prior-knowledge sheet is not available yet.\n")
}
```

# 6. Prior-Knowledge Descriptives

```{r prior-knowledge-descriptives}
if (coded_prior_knowledge_available) {
  prior_knowledge_summary <- prior_knowledge_scores |>
    group_by(group, lecture_type, practice_type) |>
    summarise(
      n_participants = n(),
      M_prior_knowledge = mean(prior_knowledge_score, na.rm = TRUE),
      SD_prior_knowledge = sd(prior_knowledge_score, na.rm = TRUE),
      M_correlation_definition = mean(correlation_definition_score, na.rm = TRUE),
      M_slope_definition = mean(slope_definition_score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(group)

  prior_knowledge_summary
} else {
  cat("Skipped: coded prior-knowledge sheet is not available yet.\n")
}
```

```{r plot-prior-knowledge-distribution}
if (coded_prior_knowledge_available) {
  p_prior_knowledge_distribution <- ggplot(
    prior_knowledge_scores,
    aes(x = group, y = prior_knowledge_score, fill = group)
  ) +
    geom_boxplot(
      alpha = 0.55,
      width = 0.55,
      outlier.shape = NA
    ) +
    geom_jitter(
      aes(color = group),
      width = 0.12,
      height = 0,
      alpha = 0.65,
      size = 1.9
    ) +
    stat_summary(
      fun = mean,
      geom = "point",
      shape = 23,
      size = 3,
      fill = "white",
      color = "black"
    ) +
    scale_fill_manual(values = GROUP_COLORS) +
    scale_color_manual(values = GROUP_COLORS) +
    scale_y_continuous(
      limits = c(0, 2),
      breaks = seq(0, 2, 0.5)
    ) +
    labs(
      title = "Prior Knowledge Score by Condition",
      subtitle = "Each point is one participant; white diamond shows group mean.",
      x = "Condition",
      y = "Prior knowledge score"
    ) +
    theme(
      legend.position = "none"
    )

  p_prior_knowledge_distribution

  save_plot_all(
    p_prior_knowledge_distribution,
    "prior_knowledge_distribution_by_condition",
    width = 7,
    height = 5
  )
} else {
  cat("Skipped: coded prior-knowledge sheet is not available yet.\n")
}
```

# 7. Save Prior-Knowledge Outputs

```{r save-prior-knowledge-outputs}
if (coded_prior_knowledge_available) {
  write_csv(
    prior_knowledge_scores,
    file.path(data_derived_dir, "prior_knowledge_scores.csv")
  )

  write_csv(
    prior_knowledge_summary,
    file.path(output_dir, "prior_knowledge_summary.csv")
  )

  cat("Saved prior_knowledge_scores.csv to:", data_derived_dir, "\n")
  cat("Saved prior_knowledge_summary.csv to:", output_dir, "\n")
} else {
  cat("No prior-knowledge scores saved because coded file is not available yet.\n")
}
```
