---
title: "Interest Survey Analysis"
author: "Jiarong Feng"
date: "2026-06-04"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.2,
  dpi = 300
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
})

has_kableExtra <- requireNamespace("kableExtra", quietly = TRUE)
has_emmeans <- requireNamespace("emmeans", quietly = TRUE)
has_psych <- requireNamespace("psych", quietly = TRUE)

pretty_table <- function(x, caption = NULL, digits = 3) {
  out <- knitr::kable(x, digits = digits, caption = caption, format = "html")
  if (has_kableExtra) {
    out <- kableExtra::kable_styling(
      out,
      full_width = FALSE,
      bootstrap_options = c("striped", "hover", "condensed")
    )
  }
  knitr::asis_output(out)
}

p_fmt <- function(p) {
  ifelse(
    is.na(p), NA_character_,
    ifelse(p < .001, "< .001", sprintf("%.3f", p))
  )
}

find_first_existing <- function(paths) {
  paths[file.exists(paths)][1]
}

find_first_dir <- function(paths) {
  paths[dir.exists(paths)][1]
}

# This Rmd usually lives in scripts/. The derived data folder is one level up.
data_derived_dir <- find_first_dir(c(
  file.path("..", "data", "derived"),
  file.path("data", "derived"),
  "data/derived"
))

fig_dir <- find_first_dir(c(
  file.path("..", "figures"),
  "figures"
))
if (is.na(fig_dir)) {
  fig_dir <- file.path("..", "figures")
  dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
}

GROUP_LEVELS <- c("CC", "CP", "PC", "PP")
GROUP_LABELS <- c(
  "CC" = "Control lecture + Control practice",
  "CP" = "Control lecture + PlotWise practice",
  "PC" = "PlotWise lecture + Control practice",
  "PP" = "PlotWise lecture + PlotWise practice"
)

GROUP_COLORS <- c(
  "CC" = "#CC79A7",
  "CP" = "#0072B2",
  "PC" = "#D55E00",
  "PP" = "#009E73"
)

SCALE_LEVELS <- c("TSI", "MSI", "FutureInterest")
SCALE_LABELS <- c(
  "TSI" = "Triggered interest",
  "MSI" = "Maintained interest",
  "FutureInterest" = "Future interest"
)

clean_scale_type <- function(x) {
  case_when(
    x %in% c("TSI", "triggeredSituationalInterest", "TriggeredInterest", "Triggered interest") ~ "TSI",
    x %in% c("MSI", "maintainedSituationalInterest", "MaintainedInterest", "Maintained interest") ~ "MSI",
    x %in% c("FutureInterest", "futureInterestSurvey", "Future interest") ~ "FutureInterest",
    TRUE ~ as.character(x)
  )
}

# Interest responses use a 1--6 scale. Higher should always mean higher interest.
# The two TSI boring items need to be reverse-coded. The function below marks
# boring items from any available text/ID fields so the scoring does not depend
# on one exact column name.
detect_reverse_interest_item <- function(data) {
  text_cols <- intersect(
    c("responseId", "question_label", "question", "responsePrompt", "prompt", "item_text", "answer_label"),
    names(data)
  )

  if (length(text_cols) == 0) {
    return(rep(FALSE, nrow(data)))
  }

  item_text <- data |>
    mutate(.row_id = row_number()) |>
    unite(".item_text", all_of(text_cols), sep = " ", remove = FALSE, na.rm = TRUE) |>
    pull(.item_text) |>
    str_to_lower()

  str_detect(item_text, "boring")
}

theme_report <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "none",
      strip.text = element_text(face = "bold"),
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(color = "gray35"),
      axis.title = element_text(face = "bold")
    )
}
```

# Overview

This report focuses on the interest survey outcomes collected after the assessment. 

The survey is summarized into three scales:

- **Triggered situational interest (TSI)**: immediate interest/enjoyment during the session.
- **Maintained situational interest (MSI)**: interest in correlation, slope, and statistics after the activity.
- **Future interest**: willingness to try similar studies in the future.

All scales are scored so that higher values mean higher interest. The two negatively worded TSI items about the video/session being boring are reverse-coded before the scale composites are computed. 


# Load interest files

```{r load-interest}
if (is.na(data_derived_dir)) {
  stop("Could not find data/derived. Check the knit working directory.")
}

interest_long_path <- find_first_existing(c(
  file.path(data_derived_dir, "interest_survey_long.csv"),
  file.path(data_derived_dir, "scored_interest_item_level.csv")
))

interest_wide_path <- find_first_existing(c(
  file.path(data_derived_dir, "interest_survey_wide.csv"),
  file.path(data_derived_dir, "scored_interest_participant_level.csv")
))

condition_index_path <- find_first_existing(c(
  file.path(data_derived_dir, "participant_condition_index.csv"),
  file.path(data_derived_dir, "participant_outcomes.csv")
))

if (is.na(interest_long_path)) {
  stop("Could not find interest_survey_long.csv or scored_interest_item_level.csv in data/derived.")
}

df_interest_raw <- readr::read_csv(interest_long_path, show_col_types = FALSE)

condition_index <- NULL
if (!is.na(condition_index_path)) {
  condition_index <- readr::read_csv(condition_index_path, show_col_types = FALSE) |>
    select(any_of(c("participantId", "group", "lecture_type", "practice_type"))) |>
    distinct()
}
```

```{r normalize-interest}
df_interest <- df_interest_raw

# Add condition columns if the interest file does not already contain them.
if (!all(c("group", "lecture_type", "practice_type") %in% names(df_interest)) &&
    !is.null(condition_index)) {
  df_interest <- df_interest |>
    left_join(condition_index, by = "participantId")
}

# If scale_type is absent but trialId is present, infer it from the trial.
if (!"scale_type" %in% names(df_interest) && "trialId" %in% names(df_interest)) {
  df_interest <- df_interest |>
    mutate(scale_type = clean_scale_type(trialId))
}

# Build a numeric raw answer if needed.
if (!"answer_num" %in% names(df_interest) && "answer" %in% names(df_interest)) {
  df_interest <- df_interest |>
    mutate(answer_num = suppressWarnings(as.numeric(answer)))
}

# Reverse-code negatively worded interest items.
# For the interest survey, these are the TSI items that say the video/session was boring.
# On a 1--6 scale, reverse-scored value = 7 - raw value.
if (!"reverse" %in% names(df_interest)) {
  df_interest <- df_interest |>
    mutate(reverse = detect_reverse_interest_item(cur_data_all()))
} else {
  df_interest <- df_interest |>
    mutate(
      reverse = as.logical(reverse) | detect_reverse_interest_item(cur_data_all())
    )
}

# Always construct answer_scored here so the reverse-coding rule is explicit.
# If the input file already contained answer_scored, this overwrites it to keep
# scoring consistent with the rules documented in this report.
df_interest <- df_interest |>
  mutate(answer_scored = if_else(reverse, 7 - answer_num, answer_num))

# If question labels are absent, use responseId as the label.
if (!"question_label" %in% names(df_interest) && "responseId" %in% names(df_interest)) {
  df_interest <- df_interest |>
    mutate(question_label = responseId)
}

required_cols <- c("participantId", "group", "lecture_type", "practice_type", "scale_type", "answer_scored")
missing_cols <- setdiff(required_cols, names(df_interest))
if (length(missing_cols) > 0) {
  stop(paste("Interest file is missing required columns:", paste(missing_cols, collapse = ", ")))
}

df_interest <- df_interest |>
  mutate(
    group = factor(group, levels = GROUP_LEVELS),
    lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
    practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
    scale_type = factor(clean_scale_type(scale_type), levels = SCALE_LEVELS),
    scale_label = recode(as.character(scale_type), !!!SCALE_LABELS),
    question_label = if ("question_label" %in% names(df_interest)) question_label else as.character(responseId)
  ) |>
  filter(!is.na(group), !is.na(scale_type), !is.na(answer_scored))
```

```{r data-source-note, results='asis'}
cat(
  "<p>Loaded interest item-level data from <code>",
  basename(interest_long_path),
  "</code> in <code>",
  data_derived_dir,
  "</code>.</p>",
  sep = ""
)
```


# Data checks

```{r interest-data-check}
interest_check <- df_interest |>
  group_by(group, lecture_type, practice_type, scale_type) |>
  summarise(
    n_participants = n_distinct(participantId),
    n_item_rows = n(),
    n_valid_responses = sum(!is.na(answer_scored)),
    min_score = min(answer_scored, na.rm = TRUE),
    max_score = max(answer_scored, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(scale_type = recode(as.character(scale_type), !!!SCALE_LABELS)) |>
  arrange(scale_type, group)

pretty_table(
  interest_check,
  caption = "Interest item-level data check"
)
```

```{r data-check-interpretation, results='asis'}
check_n <- interest_check |>
  group_by(scale_type) |>
  summarise(
    min_n = min(n_participants),
    max_n = max(n_participants),
    .groups = "drop"
  )

if (all(check_n$min_n == check_n$max_n)) {
  cat("<div class='note-box'><strong>Data check.</strong> Each condition has the same number of participants for each interest scale. This makes the descriptive condition comparisons easier to read.</div>")
} else {
  cat("<div class='note-box'><strong>Data check.</strong> Participant counts differ slightly across conditions or scales, so the model-estimated means are more useful than raw condition means alone.</div>")
}
```

# Build participant-level scale scores

Each participant gets one composite score per interest scale. The composite is the mean of the scored item responses for that scale.

```{r participant-level-scores}
interest_ppt <- df_interest |>
  group_by(participantId, group, lecture_type, practice_type, scale_type) |>
  summarise(
    n_items = n(),
    composite = mean(answer_scored, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    group = factor(group, levels = GROUP_LEVELS),
    scale_type = factor(scale_type, levels = SCALE_LEVELS),
    scale_label = recode(as.character(scale_type), !!!SCALE_LABELS)
  )

interest_summary <- interest_ppt |>
  group_by(scale_type, group, lecture_type, practice_type) |>
  summarise(
    n_participants = n_distinct(participantId),
    mean_interest = mean(composite, na.rm = TRUE),
    sd_interest = sd(composite, na.rm = TRUE),
    se_interest = sd_interest / sqrt(n_participants),
    median_interest = median(composite, na.rm = TRUE),
    min_interest = min(composite, na.rm = TRUE),
    max_interest = max(composite, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(scale_type = recode(as.character(scale_type), !!!SCALE_LABELS)) |>
  arrange(scale_type, group)

pretty_table(
  interest_summary,
  caption = "Participant-level composite interest scores by condition"
)
```

```{r participant-plot}
interest_plot_data <- interest_ppt |>
  mutate(
    group_label = factor(group, levels = GROUP_LEVELS),
    scale_label = factor(scale_label, levels = SCALE_LABELS)
  )

p_interest_distribution <- ggplot(
  interest_plot_data,
  aes(x = group_label, y = composite)
) +
  geom_boxplot(aes(fill = group_label), width = 0.52, alpha = 0.18, outlier.shape = NA) +
  geom_jitter(aes(color = group_label), width = 0.10, alpha = 0.55, size = 1.8) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3.0,
    fill = "white",
    color = "black"
  ) +
  facet_wrap(~ scale_label, nrow = 1) +
  scale_color_manual(values = GROUP_COLORS) +
  scale_fill_manual(values = GROUP_COLORS) +
  scale_y_continuous(limits = c(1, 6), breaks = 1:6) +
  labs(
    title = "Interest composite scores by condition",
    subtitle = "Small points are participants; white diamonds are condition means.",
    x = "Condition",
    y = "Composite interest score"
  ) +
  theme_report() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

p_interest_distribution
ggsave(file.path(fig_dir, "interest_composite_scores_by_condition.png"), p_interest_distribution, width = 9, height = 5.2, dpi = 300)
```

```{r descriptive-interpretation, results='asis'}
best_by_scale <- interest_summary |>
  group_by(scale_type) |>
  slice_max(mean_interest, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(scale_type, best_group = group, best_mean = mean_interest)

worst_by_scale <- interest_summary |>
  group_by(scale_type) |>
  slice_min(mean_interest, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(scale_type, worst_group = group, worst_mean = mean_interest)

desc_join <- best_by_scale |>
  left_join(worst_by_scale, by = "scale_type") |>
  mutate(range = best_mean - worst_mean)

desc_sentences <- desc_join |>
  mutate(
    sentence = paste0(
      "<li><strong>", scale_type, ":</strong> ", best_group,
      " is highest (M = ", sprintf("%.2f", best_mean),
      ") and ", worst_group,
      " is lowest (M = ", sprintf("%.2f", worst_mean),
      "), a raw spread of about ", sprintf("%.2f", range), " points.</li>"
    )
  ) |>
  pull(sentence)

cat("<div class='note-box'><strong>Descriptive pattern.</strong><ul>")
cat(paste(desc_sentences, collapse = "\n"))
cat("</ul>Across the three scales, CC is consistently the lowest or near the lowest condition. The PlotWise conditions look a little higher descriptively, especially for maintained and future interest, but the spread of participant scores within each condition is large.</div>")
```


# Item-level descriptives

I keep this section short. The participant-level composites are easier to use for condition comparisons, but item-level summaries help check whether one item is behaving oddly.

```{r item-level-summary}
interest_item_summary <- df_interest |>
  group_by(scale_type, responseId, question_label, group) |>
  summarise(
    n_responses = n(),
    mean_score = mean(answer_scored, na.rm = TRUE),
    sd_score = sd(answer_scored, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(scale_type = recode(as.character(scale_type), !!!SCALE_LABELS)) |>
  arrange(scale_type, responseId, group)

# Compact table: item means collapsed across condition, just to scan item difficulty.
interest_item_overall <- df_interest |>
  group_by(scale_type, responseId, question_label) |>
  summarise(
    n_responses = n(),
    mean_score = mean(answer_scored, na.rm = TRUE),
    sd_score = sd(answer_scored, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(scale_type = recode(as.character(scale_type), !!!SCALE_LABELS)) |>
  arrange(scale_type, mean_score)

pretty_table(
  interest_item_overall,
  caption = "Item-level interest scores, collapsed across conditions"
)
```

```{r item-plot}
item_plot_data <- interest_item_summary |>
  mutate(
    group = factor(group, levels = GROUP_LEVELS),
    question_label = fct_reorder(question_label, mean_score, .fun = mean)
  )

p_item_means <- ggplot(
  item_plot_data,
  aes(x = group, y = question_label, fill = mean_score)
) +
  geom_tile(color = "white", linewidth = 0.7) +
  geom_text(
    aes(label = round(mean_score, 1)),
    size = 3.2,
    color = "black"
  ) +
  facet_wrap(~ scale_type, scales = "free_y") +
  scale_fill_gradient(
    limits = c(1, 6),
    low = "#F7FBFF",
    high = "#08519C",
    breaks = 1:6
  ) +
  labs(
    title = "Item-level means by condition",
    subtitle = "Darker cells indicate higher interest ratings; numbers show item means.",
    x = "Condition",
    y = NULL,
    fill = "Mean"
  ) +
  theme_report(base_size = 11) +
  theme(
    legend.position = "right",
    panel.grid = element_blank()
  )

p_item_means

ggsave(
  file.path(fig_dir, "interest_item_means_by_condition.png"),
  p_item_means,
  width = 9,
  height = 6.5,
  dpi = 300
)
```

# Separate models by interest scale

The main model is run separately for each interest scale:

```text
interest composite ~ lecture_type * practice_type
```

This keeps the interpretation simple. The interaction term asks whether the effect of PlotWise practice depends on whether the lecture was also PlotWise.

```{r fit-interest-models}
fit_interest_model <- function(scale_name) {
  dat <- interest_ppt |>
    filter(scale_type == scale_name) |>
    mutate(
      lecture_type = factor(lecture_type, levels = c("Control", "PlotWise")),
      practice_type = factor(practice_type, levels = c("Control", "PlotWise")),
      group = factor(group, levels = GROUP_LEVELS)
    )

  lm(composite ~ lecture_type * practice_type, data = dat)
}

interest_models <- setNames(
  lapply(SCALE_LEVELS, fit_interest_model),
  SCALE_LEVELS
)

tidy_lm <- function(model) {
  coef_tbl <- as.data.frame(coef(summary(model)))
  coef_tbl$term <- rownames(coef_tbl)
  rownames(coef_tbl) <- NULL

  coef_tbl |>
    transmute(
      Term = recode(
        term,
        "(Intercept)" = "Intercept",
        "lecture_typePlotWise" = "PlotWise lecture",
        "practice_typePlotWise" = "PlotWise practice",
        "lecture_typePlotWise:practice_typePlotWise" = "PlotWise lecture × PlotWise practice",
        .default = term
      ),
      Estimate = Estimate,
      SE = `Std. Error`,
      `Test statistic` = `t value`,
      `p-value` = p_fmt(`Pr(>|t|)`)
    )
}

condition_estimates <- function(scale_name) {
  model <- interest_models[[scale_name]]

  if (has_emmeans) {
    emm <- emmeans::emmeans(model, ~ lecture_type * practice_type)
    as.data.frame(emm) |>
      mutate(
        group = case_when(
          lecture_type == "Control" & practice_type == "Control" ~ "CC",
          lecture_type == "Control" & practice_type == "PlotWise" ~ "CP",
          lecture_type == "PlotWise" & practice_type == "Control" ~ "PC",
          lecture_type == "PlotWise" & practice_type == "PlotWise" ~ "PP"
        )
      ) |>
      transmute(
        group = factor(group, levels = GROUP_LEVELS),
        lecture_type,
        practice_type,
        estimated_mean = emmean,
        SE = SE,
        lower.CL = lower.CL,
        upper.CL = upper.CL
      )
  } else {
    interest_ppt |>
      filter(scale_type == scale_name) |>
      group_by(group, lecture_type, practice_type) |>
      summarise(
        estimated_mean = mean(composite, na.rm = TRUE),
        SE = sd(composite, na.rm = TRUE) / sqrt(n()),
        lower.CL = estimated_mean - qt(.975, n() - 1) * SE,
        upper.CL = estimated_mean + qt(.975, n() - 1) * SE,
        .groups = "drop"
      )
  }
}
```

## Triggered interest model

```{r model-tsi}
pretty_table(
  condition_estimates("TSI"),
  caption = "Model-estimated condition means: triggered interest"
)

pretty_table(
  tidy_lm(interest_models[["TSI"]]),
  caption = "Model results: triggered interest"
)
```

```{r model-tsi-interpretation, results='asis'}
tsi_coef <- tidy_lm(interest_models[["TSI"]])
tsi_int_p <- tsi_coef |> filter(Term == "PlotWise lecture × PlotWise practice") |> pull(`p-value`)
tsi_est <- condition_estimates("TSI") |> arrange(desc(estimated_mean))

cat(
  "<div class='note-box'><strong>Triggered interest.</strong> ",
  "Triggered interest is highest in ", as.character(tsi_est$group[1]), " and lowest in ",
  as.character(tsi_est$group[nrow(tsi_est)]), ". ",
  "The pattern points in the expected direction for PlotWise practice, but the interaction is not clear (p = ", tsi_int_p, "). ",
  "I would describe this as a small descriptive engagement difference, not as strong evidence that the condition changed immediate interest.</div>",
  sep = ""
)
```

## Maintained interest model

```{r model-msi}
pretty_table(
  condition_estimates("MSI"),
  caption = "Model-estimated condition means: maintained interest"
)

pretty_table(
  tidy_lm(interest_models[["MSI"]]),
  caption = "Model results: maintained interest"
)
```

```{r model-msi-interpretation, results='asis'}
msi_coef <- tidy_lm(interest_models[["MSI"]])
msi_int_p <- msi_coef |> filter(Term == "PlotWise lecture × PlotWise practice") |> pull(`p-value`)
msi_est <- condition_estimates("MSI") |> arrange(desc(estimated_mean))

cat(
  "<div class='note-box'><strong>Maintained interest.</strong> ",
  "The condition ordering is ", paste(as.character(msi_est$group), collapse = " > "), ". ",
  "This is the scale where the PlotWise conditions look most consistently above CC, but the model still does not give a clean condition effect (interaction p = ", msi_int_p, "). ",
  "I would phrase this as a possible descriptive lift in maintained interest, especially relative to CC.</div>",
  sep = ""
)
```

## Future interest model

```{r model-future}
pretty_table(
  condition_estimates("FutureInterest"),
  caption = "Model-estimated condition means: future interest"
)

pretty_table(
  tidy_lm(interest_models[["FutureInterest"]]),
  caption = "Model results: future interest"
)
```

```{r model-future-interpretation, results='asis'}
future_coef <- tidy_lm(interest_models[["FutureInterest"]])
future_int_p <- future_coef |> filter(Term == "PlotWise lecture × PlotWise practice") |> pull(`p-value`)
future_est <- condition_estimates("FutureInterest") |> arrange(desc(estimated_mean))

cat(
  "<div class='note-box'><strong>Future interest.</strong> ",
  "Future interest is highest in ", as.character(future_est$group[1]), " and lowest in ",
  as.character(future_est$group[nrow(future_est)]), ". ",
  "The PlotWise lecture side is descriptively higher than the control lecture side, but this remains a tendency rather than a clear effect (interaction p = ", future_int_p, "). ",
  sep = ""
)
```

```{r model-estimate-plot}
all_estimates <- map_dfr(
  SCALE_LEVELS,
  ~ condition_estimates(.x) |> mutate(scale_type = .x)
) |>
  mutate(
    scale_label = factor(recode(scale_type, !!!SCALE_LABELS), levels = SCALE_LABELS),
    group = factor(group, levels = GROUP_LEVELS)
  )

p_model_estimates <- ggplot(
  all_estimates,
  aes(x = group, y = estimated_mean, color = group)
) +
  geom_point(size = 2.8) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.12) +
  facet_wrap(~ scale_label, nrow = 1) +
  scale_color_manual(values = GROUP_COLORS) +
  scale_y_continuous(limits = c(1, 6), breaks = 1:6) +
  labs(
    title = "Model-estimated interest means",
    subtitle = "Points show estimated condition means; intervals show 95% CIs.",
    x = "Condition",
    y = "Estimated interest score"
  ) +
  theme_report()

p_model_estimates
ggsave(file.path(fig_dir, "interest_model_estimated_means.png"), p_model_estimates, width = 9, height = 5.2, dpi = 300)
```

# Pairwise condition comparisons

These comparisons are mainly for locating the pattern among the four conditions. I would not over-read individual pairwise tests unless a difference is both clear in the plot and reasonably stable in the table.

```{r pairwise-comparisons}
pairwise_for_scale <- function(scale_name) {
  dat <- interest_ppt |>
    filter(scale_type == scale_name) |>
    mutate(group = factor(group, levels = GROUP_LEVELS))

  m <- lm(composite ~ group, data = dat)

  if (!has_emmeans) {
    return(tibble(scale_type = scale_name, note = "emmeans not installed"))
  }

  emm <- emmeans::emmeans(m, ~ group)
  as.data.frame(emmeans::contrast(emm, method = "pairwise", adjust = "tukey")) |>
    transmute(
      scale_type = scale_name,
      contrast = contrast,
      estimate = estimate,
      SE = SE,
      df = df,
      t_ratio = t.ratio,
      p_value = p.value,
      p = p_fmt(p.value)
    )
}

interest_pairwise <- map_dfr(SCALE_LEVELS, pairwise_for_scale) |>
  mutate(scale_type = recode(scale_type, !!!SCALE_LABELS))

pretty_table(
  interest_pairwise |>
    select(scale_type, contrast, estimate, SE, df, t_ratio, p),
  caption = "All pairwise condition comparisons by interest scale"
)
```

```{r pairwise-plot}
if (has_emmeans) {
  p_pairwise <- interest_pairwise |>
    mutate(
      scale_type = factor(scale_type, levels = SCALE_LABELS),
      contrast = fct_rev(factor(contrast))
    ) |>
    ggplot(aes(x = estimate, y = contrast)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(size = 2.2) +
    geom_errorbarh(aes(xmin = estimate - 1.96 * SE, xmax = estimate + 1.96 * SE), height = 0.15) +
    facet_wrap(~ scale_type, scales = "free_y") +
    labs(
      title = "Pairwise condition differences in interest",
      subtitle = "Positive estimates favor the first condition in the contrast.",
      x = "Estimated mean difference",
      y = NULL
    ) +
    theme_report()

  p_pairwise
  ggsave(file.path(fig_dir, "interest_pairwise_comparisons.png"), p_pairwise, width = 9, height = 6, dpi = 300)
}
```

# Design-driven contrasts

These contrasts are easier to connect back to the 2 × 2 design.

```{r design-contrasts}
design_contrasts_for_scale <- function(scale_name) {
  dat <- interest_ppt |>
    filter(scale_type == scale_name) |>
    mutate(group = factor(group, levels = GROUP_LEVELS))

  m <- lm(composite ~ group, data = dat)

  if (!has_emmeans) {
    return(tibble(scale_type = scale_name, note = "emmeans not installed"))
  }

  emm <- emmeans::emmeans(m, ~ group)

  contrast_list <- list(
    "PlotWise practice vs control practice" = c(CC = -0.5, CP = 0.5, PC = -0.5, PP = 0.5),
    "PlotWise lecture vs control lecture" = c(CC = -0.5, CP = -0.5, PC = 0.5, PP = 0.5),
    "PP vs all other conditions" = c(CC = -1/3, CP = -1/3, PC = -1/3, PP = 1),
    "Any PlotWise exposure vs CC" = c(CC = -1, CP = 1/3, PC = 1/3, PP = 1/3),
    "CP vs PC" = c(CC = 0, CP = 1, PC = -1, PP = 0)
  )

  as.data.frame(emmeans::contrast(emm, contrast_list, adjust = "none")) |>
    transmute(
      scale_type = scale_name,
      contrast = contrast,
      estimate = estimate,
      SE = SE,
      df = df,
      t_ratio = t.ratio,
      p_value = p.value,
      p = p_fmt(p.value)
    )
}

interest_design_contrasts <- map_dfr(SCALE_LEVELS, design_contrasts_for_scale) |>
  mutate(scale_type = recode(scale_type, !!!SCALE_LABELS))

pretty_table(
  interest_design_contrasts |>
    select(scale_type, contrast, estimate, SE, df, t_ratio, p),
  caption = "Design-driven contrasts by interest scale"
)
```

```{r design-contrast-interpretation, results='asis'}
if (has_emmeans) {
  low_p <- interest_design_contrasts |>
    filter(!is.na(p_value), p_value < .10) |>
    arrange(p_value)

  if (nrow(low_p) == 0) {
    cat("<div class='note-box'><strong>Design-contrast takeaway.</strong> None of the design-driven contrasts are especially strong. </div>")
  } else {
    lines <- low_p |>
      mutate(
        line = paste0(
          "<li><strong>", scale_type, ":</strong> ", contrast,
          " (estimate = ", sprintf("%.2f", estimate),
          ", p = ", p_fmt(p_value), ").</li>"
        )
      ) |>
      pull(line)

    cat("<div class='note-box'><strong>Design-contrast takeaway.</strong> The clearest contrasts to mention are:<ul>")
    cat(paste(lines, collapse = "\n"))
    cat("</ul>I would still keep the wording cautious, because these are survey outcomes and the participant-level spread is visible in the plots.</div>")
  }
}
```

```{r design-contrast-plot}
if (has_emmeans) {
  p_design <- interest_design_contrasts |>
    mutate(
      scale_type = factor(scale_type, levels = SCALE_LABELS),
      contrast = fct_rev(factor(contrast))
    ) |>
    ggplot(aes(x = estimate, y = contrast)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(size = 2.2) +
    geom_errorbarh(aes(xmin = estimate - 1.96 * SE, xmax = estimate + 1.96 * SE), height = 0.15) +
    facet_wrap(~ scale_type, scales = "free_y") +
    labs(
      title = "Design-driven interest contrasts",
      subtitle = "Positive estimates favor the PlotWise/first-listed side of each contrast.",
      x = "Estimated mean difference",
      y = NULL
    ) +
    theme_report()

  p_design
  ggsave(file.path(fig_dir, "interest_design_driven_contrasts.png"), p_design, width = 9, height = 6, dpi = 300)
}
```

# Overall interest score

This is optional, but useful if we want one broad interest measure. I treat it as descriptive because the three scales are conceptually different.

```{r overall-interest}
interest_overall <- interest_ppt |>
  group_by(participantId, group, lecture_type, practice_type) |>
  summarise(
    overall_interest = mean(composite, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(group = factor(group, levels = GROUP_LEVELS))

overall_summary <- interest_overall |>
  group_by(group, lecture_type, practice_type) |>
  summarise(
    n_participants = n_distinct(participantId),
    mean_overall_interest = mean(overall_interest, na.rm = TRUE),
    sd_overall_interest = sd(overall_interest, na.rm = TRUE),
    median_overall_interest = median(overall_interest, na.rm = TRUE),
    .groups = "drop"
  )

pretty_table(
  overall_summary,
  caption = "Overall interest score by condition"
)

m_overall_interest <- lm(overall_interest ~ lecture_type * practice_type, data = interest_overall)
pretty_table(
  tidy_lm(m_overall_interest),
  caption = "Model results: overall interest"
)
```

```{r overall-interest-plot}
p_overall <- ggplot(
  interest_overall,
  aes(x = group, y = overall_interest)
) +
  geom_boxplot(aes(fill = group), width = 0.55, alpha = 0.18, outlier.shape = NA) +
  geom_jitter(aes(color = group), width = 0.10, alpha = 0.55, size = 1.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white", color = "black") +
  scale_color_manual(values = GROUP_COLORS) +
  scale_fill_manual(values = GROUP_COLORS) +
  scale_y_continuous(limits = c(1, 6), breaks = 1:6) +
  labs(
    title = "Overall interest by condition",
    subtitle = "Overall interest averages the three interest scale composites.",
    x = "Condition",
    y = "Overall interest score"
  ) +
  theme_report()

p_overall
ggsave(file.path(fig_dir, "interest_overall_by_condition.png"), p_overall, width = 7, height = 5, dpi = 300)
```

# Working summary

The interest outcomes are generally positive, but they do not show a strong condition effect. The overall score is highest for PC (M = 4.62) and lowest for CC (M = 4.13). The overall lecture × practice interaction is p = 0.342.

PlotWise exposure appears descriptively associated with slightly higher interest than the all-control condition, especially for maintained interest
